---
title: "DataScienceModule"
author: "Olivia Crawley"
date: "2023-05-02"
output: html_document
---

#### 1. First we download all the essential packages. ####

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

install.packages("ggplot2")
library(ggplot2)


```


```{r}
install.packages("rlang")
```



```{r}

install.packages("dplyr")


library(dplyr)
library(rlang)
```

#### 2. Understanding the data ####

```{r}
mpg #Shows you the data 
?mpg #Question mark gives you more information about the data.
```
#### 3. Creating our first ggplot graph ####

```{r}
ggplot(data=mpg) + #Makes the coordinate system but no points -> ggplot statement
  geom_point(mapping = aes(x= displ, y = hwy)) #Adds points/aesthetics(aes) -> geom statement

#A geom dictates to ggplot how the points are plotted.

#Results: as you engine displacement increases, the less fuel efficent that car becomes.
```

#### 4. Understanding the grammer of graphics ####

## 4.1. Graphing template

```{r}

#ggplot(data= <DATA>) +
#  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

## 4.2. Changing the aesthetic points

# 4.2.1. Changing point colour by class

```{r}

ggplot(data=mpg) +
  geom_point(mapping = aes(x= displ, y = hwy, colour = class))

#changes the colour points of different variables
```

# 4.2.2. Changing point size by class

```{r}

ggplot(data=mpg) +
  geom_point(mapping = aes(x= displ, y = hwy, size = class))

# Changes the size of the points
#> Warning: Using size for a discrete variable is not advised (looks bad).
#> Interprets the graph wrong, looks like there is a size difference between the categories where there isnt.
#> We dont put continous aesthetics with non-continuous variables
```

# 4.2.3. Changing point transparency by class (alpha). 

```{r}
# Also consider trying a simple number, like 0.3 to this aesthetic.
ggplot(data=mpg) +
  geom_point(mapping = aes(x= displ, y = hwy, alpha = class))
# Changes how transparent the points are, different categories different transparency levels
```

# 4.2.4. Changing point shape by class.

```{r}

ggplot(data=mpg) +
  geom_point(mapping = aes(x= displ, y = hwy, shape = class))

#Changes the shapes of the points.
```

# 4.2.5. You can also set these properties manually, such as by offering a number or a colour. Lets make all of our points blue.

```{r}
ggplot(data=mpg) +
  geom_point(mapping = aes(x= displ, y = hwy), colour = "blue") 

#This only changes the colour of ALL of the points
#Simply chnages teh appearance of the plot, it doesn't show anything about the nature of the variable you've plotted. Has to make snese for data visualization.

```

What happens if you map an aesthetic to something other than a variable name?

```{r}

ggplot(data=mpg) +
  geom_point(mapping = aes(x= displ, y = hwy, colour = displ < 5))

#Creates a true false scenario where disply values above 5 are false and are coloured red, where values less than 5 are considered true and are coloured blue.

```

#### 5. Troubleshooting ####

Having the + in the wrong place.

```{r}
ggplot(data=mpg) 
+ geom_point(mapping = aes(x= displ, y = hwy))

#The + should be on the top line
```

#### 6. Facet and Panel Plots ####

Breaks a complex plot into subset plots for less confusion. In ggplot we use facets to do this.

# 6.1. Facet Wrap

```{r}

ggplot(data=mpg) +
  geom_point(mapping = aes(x= displ, y = hwy)) +
  facet_wrap (~ class, nrow = 2) #We only use facet_wrap for discrete variables, the ~ symbol splits two variables apart, if using for more than 1 variables we use facet_grid().

```

# 6.2. Facet Grid

```{r}
ggplot(data=mpg) +
  geom_point(mapping = aes(x= displ, y = hwy)) +
  facet_grid (drv ~ cyl) #using the ~ to seperate the variables

```
Use a . if you do not want to facet in the rowss or column dimension:

```{r}
ggplot(data=mpg) +
  geom_point(mapping = aes(x= displ, y = hwy)) +
  facet_grid (. ~ cyl) #Removes facets in the rows, change the position of . to cahnge to coloumns.

```
## 6.3. Facet Wrap: Exercise 

What does nrow and ncol do? WHat other options control the layout of the individual panels

```{r}
?facet_wrap

#facet_wrap(
 # facets,
 # nrow = NULL,
 # ncol = NULL,
 # scales = "fixed",
 # shrink = TRUE,
 # labeller = "label_value",
 # as.table = TRUE,
 # switch = deprecated(),
 # drop = TRUE,
 # dir = "h",
 # strip.position = "top"
# )

#nrow and ncol chnages the number of rows and columns

```

#### 7. Fitting simple lines ###

Scatter plot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) #Displaying the points
```
## 7.1. Fitting a smooth line through the points

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy)) #geom_smooth creates smooth line fit rather than points

#Not the most accurate line as it doesn't use a statistical model, makes a guess. BE CAREFUL.
```
## 7.2. Changing the line type and colours.


```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv, colour = drv)) #Fitting a smooth line to different subsets of data
#drv = type of drive train (f = front wheel drive, r = rear wheel drive, 4 = 4wd)
#colour = drv changes the colours for the different drive wheel types
```
## 7.3. Increasing the complexity of the graph

Start up

```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv, colour = drv),
              show.legend = FALSE, #removes the legend for the colours
              )
```
Here we plot multiple geoms on teh single plot. All we need to do is add them together. This one is nice for showing underlying data and how it relates top the geom_smooth line.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

While this worked perfectly, it is not ideal in a programming sense and can increase the chance of errors due to duplicated rows.

Here we make the exact same graph as above but using a better programming technique.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth() #Stops us from getting errors and helps us make our own customisations. 

#We can also make graphs and then give them a name so that we can join them together. i.e. n = ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
 # geom_point() + 
 # geom_smooth()

#n + ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
 # geom_point() + 
 # geom_smooth()
```

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + #Changes the colour of the points 
  geom_smooth()

```
The same goes if you want to specifiy different data for each layer. We can use the filter (class = 'subcompact') to select a subset of the data and plot only that subset.

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

## 7.4. Exercise: ##

1. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
2. Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.
3. Will these two graphs look different? Why/why not?

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))

#Look the exact same

```
#### 8. Transformations and Stats

## 8.1. Plotting statistics

 Our first bar chart shows that more diamonds are available with igh quality cuts then low quality cuts.
 
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes ( x = cut)) #Plots a bar plot with count on the y axis and cut on the x axis. However count os not a variable in the diamonds data -> this is something ggplot creates. The algorithm used to calculate new values for a graph is called a stat, short for statistical transformation. 

```
 You can use the geoms and stats interchangeably. For example, you can recreate the previous plot using stat_count() instead of geom_bar().
 
```{r}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut)) #produces the same graph, just different function
```
 
## 8.1.1. Overriding defaults in ggplot2

You might want to override a default stat now that you understand what the default is. Change the default stat (which is a count, a summary) to identity (which is the raw value of a variable).

```{r}
#We have to understand what the defult is doing to our work.

demo <- tribble (
  ~cut,         ~freq, #Overides the default count function so you can put the values in manually.
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551,
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```
You can also override a defult mapping from trasnformed variables to aesthetics. For instance, you could display a bar chart of *proportion* of your total diamond dataset rather than a count.

```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1)) #stat(prop) changes to proportions rather than counts

#> Warning: stat(prop) was deprecated in ggplot2 3.4.0. Please use 'after_stat(prop) instead.

?group
#what does group mean

```
## 8.2. Plotting Statistical Details

We can show more about these transformations in our plots, which is good practice to be transparent about uncertainty or any other limitation of your data.

You can do this by using stat_summary()

```{r}
ggplot(data=diamonds) +
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median
  ) #no longer bar graphs.
```
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = after_stat(prop)))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = after_stat(prop)))
```

#### 9. Positional adjustments ####

Another way we can boost the way we convey information with ggplots is by using the colour or fill to change aspects of bar colour.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut)) #colour outline around bars
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut)) # bars filled with colour
```
 No we use these aesthetics to colour by another variable like clarity. Notice how the stacking is done automatically. This is done behind the scenes with a position argument.
 
```{r}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity))
```
 The ability to make position adjustments is vital, it allows you to customise your plots in three ways, identity (raw data), fill (changes heights) and dodge (which forces ggplot2 to not put things on top of each other)

If you use position = "identity", you will be able to place each object exactly where it falls in the context of the graph. This is vital for point charts like scatter plots but makes a mess in a bar plot situation by showing too much information (a bar plot generally summarises information). We will need to alter the bar aesthetic.

```{r}
#To alter transparency (alpha)
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")

#To color the bar outlines with no fill color
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")

```

- Position = "fill" works like stacking, but makes each set of stacked bars the same height.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill") #bars are the same height
```
- Position = "dodge" places overlapping objects directly beside one another.

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge") #puts them beside each other
```
One bonus position adjustment is jitter, which slightly moves points so you can see them all (especially when they overlap). I’m sure you’ve seen really nice box plots with jittered points all over them, which you handle with this function.

- position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter") #not useful for barplots
```

#### 10. The layered grammar of graphics ####

Later in this module we’ll learn about coordinates, which allow for mapping, but for now let’s keep it this simple. Most of the time you won’t need to populate all of these, because ggplot2 has defaults for most of them, but overall you have enough here to build almost any plot.

Updated Template:

```{r}
# ggplot(data = <DATA>) + 
#  <GEOM_FUNCTION>(
 #    mapping = aes(<MAPPINGS>),
 #    stat = <STAT>, 
 #    position = <POSITION>
 # ) +
  #<FACET_FUNCTION>
```

################################################# 

#### DAY TWO ####

#### 11. Adding Labels ####

## 11.1. Main Title

Creating good labels for plots to help communicate our key findings and comply with the formatting requirements of any outlet (Scientific article or company style)

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size") #Adding a main title

# We use labs() function to create titles.
```
## 11.2. Subtitles and captions

The subtitle function adds additional detail in a smaller font beneath the title. Whereas caption adds text at the bottom right of the plot.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight.", #creates subtitle
    caption = "Data from fueleconomy.gov") #creates caption

```
## 11.3. Replacing axis and legand labels

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)", #changes x axis label
    y = "Highway fuel economy (mpg)", #changes y axis label
    colour = "Car type" ) #creates legend title
  
```
#### 12. Annotations ####

## 12.1. Creating text to the plot itself

Sometimes you’ll want to do this to highlight a single observation, for instance. Here you can use geom_text() to add textual labels to your plots. This works in the same way as geom_point() but rather than a shape geometry it can add a label. 

Here we will filter the data (more to come on this in the next workshop!) and add a label that calls in the values from the data frame. This is really neat, because it allows us to pull data values out (like the most fuel efficient car!) from the data itself and include it in the plot. This is particularly nice when these might change on an update of the data. 

```{r}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class) #adds text to the plot

#text is overlapping -> there are ways we can handle this issue. For example we can use the nudge() function teh move text a certain amount and using other R packages to handle text wrapping. 



```
Labels are great and allow lots of customisation, even allowing actions such as highlighting the maximum or minimum value, which you can obtain using base R functions (allowing them to change if your data change). 

We won’t cover the ins and outs of these here, we just want you to know that you can use labels, they’re super flexible, and allow you develop professional and polished outputs.

#### 13. Scales ####

Normally ggplot adds automatic scales to graphs for you but we can change these to help with customization. 

```{r}
#Automatic scales

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) #automatically adds defult scales behind the scenes

```
```{r}
#Adjusting scales

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous("displ", 1 ) +
  scale_y_continuous("hwy", 7) +
  scale_colour_discrete("class", 4)

#scale_x_continuous(name, breaks, labels, limits, trans)
#scale_y_continuous(name, breaks, labels, limits, trans)
```
#### 14. Axis Ticks ####

You can also change the ticks on your axes. Breaks controls the position of the ticks and you can use labels as an argument to change the text label associated with ticks. 

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5)) #creates y axis ticks every 5 points between 15 and 40


```
Similarly, you can use labels set to NULL to suppress the labels altogether. This is sometimes an option if you’re trying to format your plots in a particular way. 

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL) #Removes tick labels from the x and y axis
```

#### 15. Legends and colour schemes ####

Changing the position of the legend using the theme() setting. 

```{r}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

base + theme(legend.position = "left") #moves legend to the left side of graph
base + theme(legend.position = "top") #moves legend above graph
base + theme(legend.position = "bottom") #moves legend below graph
base + theme(legend.position = "right") # the default, right of the graph
base + theme(legend.position = "none") #removes the display of the legend
```

#### 16. Replacing a scale ####

There are two types of scales you're most likely to swith out, continuous position scales and colour scales.

It’s very useful to plot transformations of your variable. For example, as we’ve seen in diamond prices it’s easier to see the precise relationship between carat and price if we log transform them using a ggplot2 function for scale precisely for this purpose.

## 16.1. Log-transform 

```{r}
ggplot(diamonds, aes(carat, price)) + #looking at carat and price
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10()
```
## 16.2.1. Colour scale customization 

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) #gives you the defult colour palette

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) + #Colours the drv differently
  scale_colour_brewer(palette = "Set1") #Changes the colour palette of the points
```
If there are just a few colours, you can add a redundant shape mapping. This will also help ensure your plot is interpretable in black and white.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) + #shape function adds a redundant shape mapping (changes the shapes of the diffretn drv points while still keeping the colour)
  scale_colour_brewer(palette = "Set1")

```
## 16.2.2. Manually selecting colour scales

```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue")) #manually sleects colours for the different variables (good if you want specific colours that represent something in your data)

#Have a look at the viridis colour scheme!
```
## 16.2.3. Viridis Colour Scheme

Here we look at the viridis colour scheme using scale_colour_viridis().

```{r}
install.packages('viridis')
install.packages('hexbin')
library(viridis)
library(hexbin)

```

```{r}
df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed() #fixed coordinates

ggplot(df, aes(x, y)) +
  geom_hex() + #a new geom
  viridis::scale_fill_viridis() + #adds the colour scale
  coord_fixed()

```

#### 17. Themes ####

Customizing the entire theme of the plot. Changes some or all of the non-data elements of your plot with a theme.

ggplot has eight themes by defults but many more can be added using add-on packages. 

HERE WE LOOK AT HOW THE GRID CHANGES.

## 17.1. Default theme

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()
```

## 17.2. Light Theme

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light()

```

## 17.3. Classic theme

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic()
```
## 17.4. Dark theme

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark()
```
## 17.5. Custom Theme

One particularly nice thing you can do though, is to develop your own themes simply by setting some or all of the arguments to theme() yourself. Here’s one I wrote so that my plots are always the same style:

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial"))
```
#### 18. Saving and exporting your plots
Well, now you’ve iteratively developed your plot, you want to save it to disk in some format that allows you to import into your report or to share it with the world. 
In general, the best way to make a single file will be to export it using the ggsave() function. 
There are two main ways to get your plots out of R using ggplot2. ggsave() will save the most recent plot to your working directory. 

```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")
#> Saving 7 x 4.32 in image
```

Sometimes, you’ll want to play with the dimensions of your plot, in which case you can use the width and height arguments to this function. Try it!
Note that you can alternatively export your results, together with your code, using R markdown via knitr. 
** If you are up to this point it is time to work on your Plot Deconstruction exercise. Please continue with this (in groups of 2) for the remainder of this workshop.**


###################### 


